---
title: "6: Physical Properties of Rive"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Compute recurrence intervals for stream discharge 
2. Analyze the effects of watershed disturbance on recurrence intervals and interpret results against the concept of stationarity
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
How is climate change impacting discharge in streams and rivers? What are the new and ongoing challenges faced by these impacts in watersheds?

## Session Set Up

```{r, message = FALSE}
getwd()

library(tidyverse)
library(dataRetrieval)
library(lubridate)


theme_set(theme_classic())
```

## Recurrence Intervals and Exceededence Probability

A **recurrence interval** is the past recurrence of an event, in this case a peak annual discharge measurement of a given magnitude. The value of a recurrence interval corresponds to the average number of years between discharge of a given magnitude. Typically the minimum amount of years required to construct a recurrence interval is 10, but 30 is more robust. A recurrence interval, T, is calculated as: 

$T = (n+1)/m$

where n is the number of years and m is the ranking of an event within the observed period. We add one to n because we are computing the recurrence interval for a discharge event of a given magnitude *or greater*. 

Similarly, we can calculate an **exceedence probability**, or the probability of encountering a discharge event of a given magnitude or greater in any given year: 

$P = 1/T$

This is where the terms "100-year flood" and similar are derived. Remember this is a probability based on past occurrence, not an accurate forecast of how often we will see that event happening. When current patterns of discharge differ from past patterns, we observe **nonstationary** behavior. Nonstationarity results in events that occur more or less frequency than predicted based on the exceedence probability.

### Has Eno River dicharge displayed stationary behavior over the period of record?

Let's import discharge data for the Eno River near Durham for all available dates.

```{r}
EnoDischarge <- readNWISdv(siteNumbers = "02085070",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "",
                     endDate = "")

names(EnoDischarge)[4:5] <- c("Discharge", "Approval.Code")

attr(EnoDischarge, "variableInfo")
attr(EnoDischarge, "siteInfo")
  
# Build a ggplot
EnoPlot <- 
  ggplot(EnoDischarge, aes(x = Date, y = Discharge)) +
         geom_line() +
         xlab("Year")
print(EnoPlot)
```

We can then compute recurrence intervals based on the first 30 years of data.

```{r}  
EnoDischarge <- 
  EnoDischarge %>%
  mutate(Year = year(Date))

EnoRecurrence <- 
  EnoDischarge %>%
  filter(Year < 1993) %>%
  group_by(Year) %>%
  summarise(PeakDischarge = max(Discharge)) %>% 
  mutate(Rank = rank(-PeakDischarge), 
         RecurrenceInterval = (length(Year) + 1)/Rank, 
         Probability = 1/RecurrenceInterval)

EnoPeakPlot <- 
  ggplot(EnoRecurrence, aes(x = Year, y = PeakDischarge)) +
         geom_bar(stat = "identity") +
         xlab("Year")
print(EnoPeakPlot)
```

Let's display and model the relationship between peak annual disharge and recurrence interval. We can use the statistical model to compute discharge for recurrence intervals that occur above the 30-year mark.
```{r}
EnoRecurrencePlot <- 
  ggplot(EnoRecurrence, aes(x = RecurrenceInterval, y = PeakDischarge)) +
  geom_point() #+
  #scale_x_log10() 
print(EnoRecurrencePlot)

Eno.RImodel <- lm(data = EnoRecurrence, PeakDischarge ~ log(RecurrenceInterval))
summary(Eno.RImodel)

#What is the discharge for a 100-year flood in this system? a 500-year flood?
Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log(100)
Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log(500)
```

What if we were to build a recurrence interval model for the entire period of record? How would this compare to the 30-year construction?
```{r}
EnoRecurrence.Full <- 
  EnoDischarge %>%
  group_by(Year) %>%
  summarise(PeakDischarge = max(Discharge)) %>% 
  mutate(Rank = rank(-PeakDischarge), 
         RecurrenceInterval = (length(Year) + 1)/Rank, 
         Probability = 1/RecurrenceInterval)

EnoRecurrencePlot.Full <- 
  ggplot(EnoRecurrence, aes(x = RecurrenceInterval, y = PeakDischarge)) +
  geom_point() +
  geom_point(data = EnoRecurrence.Full, color = "#02818a",
             aes(x = RecurrenceInterval, y = PeakDischarge))
print(EnoRecurrencePlot.Full)

Eno.RImodel.Full <- lm(data = EnoRecurrence.Full, PeakDischarge ~ log(RecurrenceInterval))
summary(Eno.RImodel.Full)
Eno.RImodel.Full$coefficients
Eno.RImodel$coefficients

Eno.RImodel.Full$coefficients[1] + Eno.RImodel.Full$coefficients[2]*log(100)
Eno.RImodel.Full$coefficients[1] + Eno.RImodel.Full$coefficients[2]*log(500)

Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log(100)
Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log(500)

```

What differences did you see for the recurrence intervals built under different periods of record? How would your prediction of flood events differ if you were to use these models for forecasting purposes? 

>

What would you recommend for a watershed manager seeking to build the most accurate recurrence interval model for the Eno River?

> 


### Examining the effects of urbanization on discharge

Salado Creek is located in San Antonio, Texas, an area that has been rapidly urbanizing over the course of the last several decades (http://worldpopulationreview.com/us-cities/san-antonio-population/#byPopulation). Using the code above, compute recurrence intervals for Salado Creek for the first 30 years of record and for the most recent 30 years of record. How do the graphs and models differ? How do your predictions of 100- and 500-year floods differ?

```{r}
# Import data
SaladoDischarge <- readNWISdv(siteNumbers = "08178700",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "",
                     endDate = "")
names(SaladoDischarge)[4:5] <- c("Discharge", "Approval.Code")
attr(SaladoDischarge, "siteInfo")
  
SaladoPlot <- 
  ggplot(SaladoDischarge, aes(x = Date, y = Discharge)) +
         geom_line() +
         xlab("Year")
print(SaladoPlot)

# add more code here: 

```

### Examining the effects of dam construction on recurrence intervals

The stream gage in the Green River near Auburn, Washington, is located directly downstream of the Howard A. Hanson Dam. The dam was built in 1961 for flood control purposes, and the reservoir now provides water supply to the city of Tacoma. How have peak discharges changed since the construction of the dam?

Using the code above, construct recurrence intervals for the periods before and after the construction of the dam. How do the graphs and models differ? How do your predictions of 100- and 500-year floods differ?

```{r}
GreenDischarge <- readNWISdv(siteNumbers = "12113000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "",
                     endDate = "")
names(GreenDischarge)[4:5] <- c("Discharge", "Approval.Code")
attr(GreenDischarge, "siteInfo")
  
GreenPlot <- 
  ggplot(GreenDischarge, aes(x = Date, y = Discharge)) +
         geom_line() +
         xlab("Year")
print(GreenPlot)

# add more code here:

```
## Closing Discussion
This week we focused on discharge as a physical property of a stream or river. How might you use your knowledge of discharge to inform other physical processes occurring in rivers?
