---
title: "Assignment 5: Water Quality in Lakes"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Hydrologic Data Analysis on water quality in lakes

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single HTML file.
8. After Knitting, submit the completed exercise (HTML file) to the dropbox in Sakai. Add your last name into the file name (e.g., "A05_Salk.html") prior to submission.

The completed exercise is due on 2 October 2019 at 9:00 am.

## Setup

1. Verify your working directory is set to the R project file, 
2. Load the tidyverse, lubridate, and LAGOSNE packages. 
3. Set your ggplot theme (can be theme_classic or something else)
4. Load the LAGOSdata database and the trophic state index csv file we created on 2019/09/27.

```{r setup}

```

## Trophic State Index

5. Similar to the trophic.class column we created in class (determined from TSI.chl values), create two additional columns in the data frame that determine trophic class from TSI.secchi and TSI.tp (call these trophic.class.secchi and trophic.class.tp). 

```{r}

```


6. How many observations fall into the four trophic state categories for the three metrics (trophic.class, trophic.class.secchi, trophic.class.tp)? Hint: `count` function.

```{r}

```

7. Find and quantify instances where trophic.class and trophic.class.secchi differ. Do the same comparing trophic.class and trophic.class.tp. 

```{r}

```

What proportion of observations do these metrics differ?

> 

Note: To take this further, a researcher might determine which trophic classes are susceptible to being differently categorized by the different metrics and whether certain metrics are prone to categorizing trophic class as more or less eutrophic. This would entail more complex code.

## Nutrient Concentrations

8. Create a data frame that includes the columns lagoslakeid, sampledate, tn, tp, state, and state_name. Mutate this data frame to include sampleyear and samplemonth columns as well. Call this data frame LAGOSNandP. 

```{r}

```

9. Create two violin plots comparing TN and TP concentrations across states. Include a 50th percentile line inside the violins.
```{r}

```

Which states have the highest and lowest median concentrations?

> TN:

> TP:

Which states have the highest and lowest concentration ranges?

> TN: 

> TP: 

10. Create two jitter plots comparing TN and TP concentrations across states, with samplemonth as the color. Choose a color palette other than the ggplot default.

```{r}

```

Which states have the most samples? How might this have impacted total ranges from #9?

> TN:

> TP: 

Which months are sampled most extensively? Does this differ among states?

> TN:

> TP: 

11. Create two jitter plots comparing TN and TP concentrations across states, with sampleyear as the color. Choose a color palette other than the ggplot default.

```{r}

```


Which years are sampled most extensively? Does this differ among states?

> TN:

> TP: 

## Reflection
12. What are 2-3 conclusions or summary points about lake water quality you learned through your analysis?

>

13. What data, visualizations, and/or models supported your conclusions from 12?

> 

14. Did hands-on data analysis impact your learning about water quality relative to a theory-based lesson? If so, how?

>

15.	How did the real-world data compare with your expectations from theory?

>
