---
title: "Assignment 7: High Frequency Data"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Hydrologic Data Analysis on high frequency data

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single pdf file.
5. After Knitting, submit the completed exercise (pdf file) to the dropbox in Sakai. Add your last name into the file name (e.g., "A07_Chamberlin.pdf") prior to submission.

The completed exercise is due on 16 October 2019 at 9:00 am.

## Setup

1. Verify your working directory is set to the R project file, 
2. Load the StreamPULSE, streamMetabolizer and tidyverse packages. 
3. Set your ggplot theme (can be theme_classic or something else)


```{r setup}

```


4. Download data from the Stream Pulse portal using `request_data()` for the Kansas River, ("KS_KANSASR"). Download the discharge (`Discharge_m3s`), disolved oxygen (`DO_mgL`) and nitrate data (`Nitrate_mgL`) for the entire period of record

5. Reformat the data into one dataframe with columns DateTime_UTC, DateTime_Solar (using `convert_UTC_to_solartime()`), SiteName, DO_mgL, Discharge_m3s, and Nitrate_mgL.
```{r Datadownload}


```

6. Plot each of the 3 variables against solar time for the period of record

```{r}

```

7. How will you address gaps in these dataseries?

>

8. How does the daily amplitude of oxygen concentration swings change over the season? What might cause this?

>

## Baseflow separation
9. Use the `EcoHydRology::BaseflowSeparation()` function to partition discharge into baseflow and quickflow, and calculate how much water was exported as baseflow and quickflow for this time period. Use the DateTime_UTC column as your timestamps in this analysis.

The `package::function()` notation being asked here is a way to call a function without loading the library. Sometimes the EcoHydRology package can mask tidyverse functions like pipes, which will cause problems for knitting. In your script, instead of just typing `BaseflowSeparation()`, you will need to include the package and two colons as well.

10. Create a ggplot showing total flow, baseflow, and quickflow together. 


```{r}

```


11. What percentage of total water exported left as baseflow and quickflow from the Kansas River over this time period?

>

12. This is a much larger river and watershed than the 2 we investigated in class. How does the size of the watershed impact how flow is partitioned into quickflow and baseflow? 

>

13. The site we are looking at is also further down in its river network (i.e. instead of being a headwater stream, this river has multiple tributaries that flow into it). How does this impact your interpretation of your results?

>

## Chemical Hysteresis

14. Create a ggplot of flow vs. nitrate for the large storm in May (~May 1 - May 20). Use color to represent Date and Time.

```{r}

```

15. Does this storm show clockwise or counterclockwise hysteresis? Was this storm a flushing or diluting storm?

>

16. What does this mean for how nitrate gets into the river from the watershed?

>

## Reflection
17. What are 2-3 conclusions or summary points about high frequency data you learned through your analysis?

>

18. What data, visualizations, and/or models supported your conclusions from 17?

> 

19. Did hands-on data analysis impact your learning about high frequency data relative to a theory-based lesson? If so, how?

>

20.	How did the real-world data compare with your expectations from theory?

>
